On Misuses of the Kolmogorov--Smirnov Test for One-Sample Goodness-of-Fit

### 2023-04-20 manuscript/manuscript.tex
mid-way through Sec 3

The null hypothesis of a goodness-of-fit test is often a composite hypothesis
instead of a single hypothesis. That is, the hypothesized distribution is a
family of distributions with unspecified parameters instead of a specific member
in this family. Let $F_\theta$ be a family of distributions indexed by parameter
vector~$\theta$. The null hypothesis is
\[
  H_0: \text{the random sample $X_1, \ldots, X_n$
    comes from a distribution $F_\theta$ for some $\theta$.}
\]
Since $\theta$ is unknown, one would naturally estimate $\theta$ by
$\hat\theta_n$ from, for example, maximum likelihood or methods of moments. The
KS statistic would then be computed as
\begin{equation}
  \label{eq:ks_fitted}
  D_n = \sqrt{n} \sup_x | F_n(x) - F_{\hat\theta_n}(x) |.
\end{equation}
An overly large $D_n$ still indicates evidence to reject~$H_0$. To get the
p-value of the observed $D_n$, however, note that the null distribution is not
the same as that in the standard
case~\eqref{eq:ks_standard}. If the same null distribution were used, one would
be testing a different~$H_0$ that the random sample comes from the specific
member distribution~$F_{\hat\theta}$ instead of the family~$F_\theta$.

The consequence of using the wrong null distribution for the KS statistic $D_n$
in~\eqref{eq:ks_fitted} can be illustrated through a simple simulation study. A
random sample $X_1, \ldots, X_n$ was generated from the standard normal
distribution with sample size $n = 100$. The sample mean and sample standard
deviation were used as $\hat\theta_n$. The p-value of the test
statistic~$D_n$ in~\eqref{eq:ks_fitted} was obtained by naively calling the
\texttt{ks.test()} function in~R with the fitted normal distribution as the
hypothesized distribution. That is, the hypothesized distribution was
$N(\bar X, s^2)$ where $\bar X$ is the sample mean and $s^2$ is the sample
variance. This experiment was repeated 1000 times and the histogram of the
1000 p-values are displayed in Figure~\ref{fig:pp_f}. If the test were
valid, the p-values would be uniformly distributed in the unit
interval. Clearly, the pattern of the 1000 p-values are very different from what
one would expect to see from 1000 draws from the standard uniform
distribution.

@@

Nonparametric bootstrap can also be used to approximate the null distribution
of the test statistic except that a bias correction is needed
\citet{babu2004goodness}. Details of the nonparametric bootstrap procedure is
in the Appendix. The results from nonparametric bootstrap are similar to those
from the parametric bootstrap and, hence, are omitted.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=.5\textwidth]{pp_f.pdf}
  \caption{PP-plot for the p-values obtained from KS tests with unspecified
    parameters with sample size $n = 200$ based on 1000 replicats. The data
    generating models were $N(8, 8)$ and $\Gamma(8, 1)$.
  }
  \label{fig:pp_f}
\end{figure}

In the same simulation study, p-values were obtained for the 1000 replicates
with both the parametric and nonparametric bootstrap procedures. The center and
right panels of Figure~\ref{fig:pp_f} display the histograms of the 1000
p-values from the parametric and nonparametric bootstrap procedures,
respectively. This time, both samples of p-values appear to come from
standard uniform distributions.

@@

The invalidity of the standard statistic for serially dependent data can be
illustrated by a simple simulation study. Consider a first-order autoregressive
or AR(1) model with marginal standard normal distribution. For each AR
coefficient $\psi \in \{-0.3, 0.2, 0.1, 0, 0.1, 0.2, 0.3\}$, we generated 1000
series of length 100. For each series, we applied the KS statistic
in~\eqref{eq:ks_standard} to test $H_0$ that the $X_i$'s follow $N(0, 1)$
marginally. The histograms of the p-values from $1000$ replicates for all the
$\psi$ values are displayed in Figure~\ref{fig:pp_s}. When
$\phi \ne 0$, the histograms do not look likely to be generated from a standard
uniform distribution. Positive $\psi$ values led to p-values with a distribution
function higher than the standard uniform distribution function and, hence,
conservative tests. Negative $\psi$ values led to p-values with a distribution
function lower than the standard uniform distribution and, hence, liberal
tests. The larger the magnitude of
$\psi$, the bigger the deviation from the desired level.

>>>>

The null hypothesis of a goodness-of-fit test is often a composite hypothesis
instead of a single hypothesis. That is, the hypothesized distribution is a
family of distributions with unspecified parameters instead of a specific member
in this family. Let $F_\theta$ be a family of distributions indexed by parameter
vector~$\theta$. The null hypothesis is
\[
  H_0: \text{the random sample $X_1, \ldots, X_n$
    comes from a distribution $F_\theta$ for some $\theta$.}
\]
Since $\theta$ is unknown, one would naturally estimate $\theta$ by
$\hat\theta_n$ from, for example, maximum likelihood or methods of moments. The
KS statistic would then be computed as
\begin{equation}
  \label{eq:ks_fitted}
  D_n = \sqrt{n} \sup_x | F_n(x) - F_{\hat\theta_n}(x) |.
\end{equation}
An overly large $D_n$ still indicates evidence to reject~$H_0$. To get the
p-value of the observed $D_n$, however, note that the null distribution is not
the same as that in the standard
case~\eqref{eq:ks_standard}. If the same null distribution were used, one would
be testing a different~$H_0$ that the random sample comes from the specific
member distribution~$F_{\hat\theta_n}$ instead of the family~$F_\theta$.

The consequence of using the wrong null distribution for the KS statistic $D_n$
in~\eqref{eq:ks_fitted} can be illustrated through a simple simulation study. A
random sample $X_1, \ldots, X_n$ was generated from a normal
distribution with both mean and variance parameters set equal to 8 ($N(8,8)$), 
and sample size $n = 200$. The sample mean and sample standard
deviation were used as $\hat\theta_n$. The p-value of the test
statistic~$D_n$ in~\eqref{eq:ks_fitted} was obtained by naively calling the
\texttt{ks.test()} function in~R with the fitted normal distribution as the
hypothesized distribution. That is, the hypothesized distribution was
$N(\bar X, s^2)$ where $\bar X$ is the sample mean and $s^2$ is the sample
variance. This experiment was repeated 1000 times and the 
probability-probability plot (PP-plot) of the 1000 p-values are displayed in the 
top left plot of Figure~\ref{fig:pp_f}. If the test were
valid, the p-values would be uniformly distributed in the unit
interval and the points in the PP plot would fall along the diagonal line. 
Clearly, the distribution of the 1000 p-values are very different from what
one would expect to see from 1000 draws from the standard uniform
distribution.

We similarly considered random samples $X_1, \ldots, X_n$ 
generated from a gamma distribution with shape and scale parameters 8 and 1, 
respectively ($\Gamma(8,1)$), and sample size $n=200$. These parameter values 
were selected so that mean and variance of the normal and gamma distributions 
in the simulations are the same. Likewise, the p-value of the test 
statistic~$D_n$ in~\eqref{eq:ks_fitted} was obtained by naively calling the
\texttt{ks.test()} function in~R with the fitted gamma distribution as the
hypothesized distribution using the shape and scale \eds{maximum likelihood 
estimates}. The PP plot of the p-values from 1000 replications of this 
experiment are displayed in the bottom left plot of Figure~\ref{fig:pp_f}. The
p-values are again non-uniformly distributed due to the use of the 
estimated parameters when specifying the null distribution.

@@

In the same simulation study, p-values were obtained for the 1000 replicates
from the parametric bootstrap procedures. The top and bottom right plots of 
Figure~\ref{fig:pp_f} display the PP-plots of the 1000
p-values from the normal and gamma data generation scenarios,
respectively. This time, all points fall along the diagonal lines implying that
the p-values are coming from standard uniform distributions.

Nonparametric bootstrap can also be used to approximate the null distribution
of the test statistic in~\eqref{eq:ks_fitted} except that a bias correction is 
needed
\citep{babu2004goodness}. Details of the nonparametric bootstrap procedure are
in the Appendix. The results from nonparametric bootstrap are similar to those
from the parametric bootstrap and, hence, are omitted.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=.5\textwidth]{pp_f.pdf}
  \caption{PP-plots for the p-values obtained from KS tests with unspecified
    parameters with sample size $n = 200$ based on 1000 replicates. The data
    generating models were $N(8, 8)$ and $\Gamma(8, 1)$.
  }
  \label{fig:pp_f}
\end{figure}

@@

The invalidity of the standard statistic for serially dependent data can be
illustrated by a simple simulation study. Consider a first-order autoregressive
or AR(1) model with marginal normal (or gamma) distribution. For each AR
coefficient $\psi \in \{-0.3, 0.1, 0.1, 0.3\}$, we generated 1000
series of length 100. For each series, we applied the KS statistic
in~\eqref{eq:ks_standard} to test $H_0$ that the $X_i$'s follow $N(8, 8)$ (or 
$\Gamma(8,1)$ marginally. The PP-plots of the p-values from $1000$ replicates 
for all the $\psi$ values are displayed in Figure~\ref{fig:pp_s}. When
$\phi$ diverges from 0, the p-values do not look likely to be generated from a 
standard uniform distributions. Positive $\psi$ values led to p-values with a 
distribution function higher than the standard uniform distribution function and, 
hence, conservative tests. Negative $\psi$ values led to p-values with a 
distributionfunction lower than the standard uniform distribution and, hence, 
liberal tests. \eds{check previous statement, not consistent with beginning of 
this section} The larger the magnitude of
$\psi$, the bigger the deviation from the desired level.
\eds{check that $\psi=.1$ plots are really from .1 and not 0}
